\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3.2\baselineskip}

### 2017-03-24 paper/ms.tex
Pedantic footnote placements.

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that are computationally tractable, even with large
datasets.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
In all of these cases, we infer posterior distributions over models for the
power spectrum of the process that generated these irregularly sampled time
series without computing a standard algebraic estimator.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

In the following pages, we motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released well tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

@@

Similarly, setting both $b_j$ and $d_j$ to zero, we get a Ornstein--Uhlenbeck
process
\begin{eqnarray}
k_j(\tau_{nm}) = a_j\,\exp\left(-c_j\,\tau_{nm}\right)
\end{eqnarray}
with the PSD
\begin{eqnarray}
S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
    \frac{1}{1+{\left(\frac{\omega}{c_j}\right)}^2} \quad.
\end{eqnarray}

@@

\item \texttt{lapack}: The general banded LU decomposition implementation from
    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
    \project{BLAS} routines\footnote{The experiments below use the Intel Math
    Kernel Library (\project{MKL}):
    \url{https://software.intel.com/en-us/intel-mkl}.}, and

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key take away here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific but flexible class of one-dimensional kernel
functions.
The closest cousin is the CARMA method with similar strengths and limitations
but, since there is a simpler relationship between the parameters of a
\celerite\ model and its behavior, \celerite\ can be more easily integrated
with physical or physically interpretable models.

@@

Alongside this paper, we have released a well tested and documented open
source software package that implements the method and all of the examples
discussed in these pages.
This software is available at \url{https://github.com/dfm/celerite}\footnote{%
This version of the paper was generated with git commit \texttt{\githash}
(\gitdate).} and it is made available under the MIT license.

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that are computationally tractable, even with large
datasets.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
In all of these cases, we infer posterior distributions over models for the
power spectrum of the process that generated these irregularly sampled time
series without computing a standard algebraic estimator.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well-tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

In the following pages, we motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released well-tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

@@

Similarly, setting both $b_j$ and $d_j$ to zero, we get an Ornstein--Uhlenbeck
process
\begin{eqnarray}
k_j(\tau_{nm}) = a_j\,\exp\left(-c_j\,\tau_{nm}\right)
\end{eqnarray}
with the PSD
\begin{eqnarray}
S_j(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{a_j}{c_j}\,
    \frac{1}{1+{\left(\frac{\omega}{c_j}\right)}^2} \quad.
\end{eqnarray}

@@

\item \texttt{lapack}: The general banded LU decomposition implementation from
    \project{LAPACK}\footnote{We use the \texttt{dgbtrf} and \texttt{dgbtrs}
    methods from \project{LAPACK}.} \citep{Anderson:1999} using optimized
    \project{BLAS} routines,\footnote{The experiments below use the Intel Math
    Kernel Library (\project{MKL}):
    \url{https://software.intel.com/en-us/intel-mkl}.} and

@@

Many other approximate methods for scaling GP inference exist \citep[see, for
example,][and references therein]{Wilson:2015a} and we make no attempt to make
our discussion exhaustive.
The key takeaway here is that \celerite\ provides an \emph{exact} method for
GP inference for a specific but flexible class of one-dimensional kernel
functions.
The closest cousin is the CARMA method with similar strengths and limitations
but, since there is a simpler relationship between the parameters of a
\celerite\ model and its behavior, \celerite\ can be more easily integrated
with physical or physically interpretable models.

@@

Alongside this paper, we have released a well-tested and documented open
source software package that implements the method and all of the examples
discussed in these pages.
This software is available at \url{https://github.com/dfm/celerite}\footnote{%
This version of the paper was generated with git commit \texttt{\githash}
(\gitdate).} and it is made available under the MIT license.
